---
title: "medfateland - Mediterranean landscape forest functioning"
output: github_document
---

```{r setup, echo = FALSE}

knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.path = "README-")
```

## Introduction

The R package **medfateland** has been designed to extend the capabilities of package **medfate** to the landscape level. It allows running the stand-level models available in medfate on points and cells within landscape. More importantly, medfateland allows considering spatial hydrological processes in forested watersheds. Hence, medfateland can be used as a tool for eco-hydrological applications.   


## Package installation

Development versions of package **medfateland** can be found at GitHub as follows (required package `devtools` should be installed/updated first):

```{r installation, eval = FALSE}
devtools::install_github("vegmod/medfateland")
```


## Documentation

+ The package installation includes a number of *vignettes* that illustrate how to run simulation models in **medfateland**.

+ The user is also advised to read *vignettes* included in package **medfate**.

+ Additional documentation of the package, including a *reference book*, can be found at https://vegmod.github.io/software/medfate. 

