---
title: "Landscape water balance"
author: "Miquel De Caceres"
date: "`r Sys.Date()`"
output: 
  rmarkdown::html_vignette:
    toc: TRUE
vignette: >
  %\VignetteIndexEntry{Landscape water balance}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignettePackage{medfateland}
  \usepackage[utf8]{inputenc}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(medfate)
library(medfateland)
```



## Simulations on a set of forest stands

Package `medfate` include functions `spwb()` or `growth()` to simulate soil water balance on a single forest stand. This section describes how to run simulations on a set of forest stands in one call. This is done using functions `spwbpoints()`, `spwbpixels()`, `spwbgrid()` and the corresponding growth functions. 

As an example, we will use function `spwbpoints()`, which simulates soil plant water balance on forests distributed in particular locations. The function takes an object of class `SpatialPointsLandscape` as input.

### Example point data
We begin by loading an example dataset of 30 forest stands:

```{r, fig.align='center', fig.width=7}
data("examplepointslandscape")
```

We can draw maps of some variables using:

```{r, fig.align='center', fig.width=7}
plot(examplepointslandscape, "basalArea")
```

### Weather inputs and local control parameters

As before, we need meteorological data, species parameters and control parameters for local simulations (which will apply to all forest stands):

```{r}
data("examplemeteo")
data("SpParamsMED")
localControl = defaultControl()
```

### Calling the simulation function

The call to `spwbpoints()` can be done as follows (here we restrict the dates for simplicity):
```{r, warning=FALSE}
dates = seq(as.Date("2001-01-01"), as.Date("2001-02-28"), by="day")
res <- spwbpoints(examplepointslandscape, SpParamsMED, examplemeteo, 
                  dates = dates, localControl = localControl,  progress = FALSE)
```

Function `spwbpoints()` first initializes model inputs by calling `forest2spwbInput()` for each forest stand described in the `SpatialPointsLandscape` object.  Then it calls function `spwb()` for each forest stand and stores the result. The fact that we used `examplemeteo` as meteorological input involves that the same weather was applied to all forest stands, but different weather could have been specified for each one (see documentation of function `spwbpoints()`).

The result of calling `spwbpoints()` is a list (whose class name is the same as the function name) with the following elements:

```{r}
names(res)
```

Element `sp` describes spatial information, vector `xlist` contains the `spwbInput` corresponding to each forest stand (which can be used in subsequent simulations), vector `resultlist` contains the output of `spwb()` function for each forest stand (i.e. its elements are objects of the S3 class `spwb`) and `summarylist` contains summaries (if requested, see below). 

### Summaries and plots

The structure of the output of `spwbpoints()` allows querying information for the simulation of any particular forest stand. For example, we may use function `plot.spwb()`, from package **medfate**, to display the simulation results on a particular plot:
```{r, fig.width=7}
plot(res$resultlist[[1]], "Evapotranspiration")
```

If we want a monthly summary of water balance for the first stand, we can use function `summary.spwb()` from package **medfate**:

```{r}
summary(res$resultlist[[1]], freq="months",FUN=sum, 
        output="WaterBalance")
```

However, a more convenient way of generating summaries is by calculating them on all forest stands, using:

```{r}
res_sum <- summary(res, summaryFunction = summary.spwb, 
                   freq="months", output="WaterBalance")
```
where the argument `summaryFunction` points to the function to be used to generate summaries and the remaining arguments are those of the summary function. The result is an object of class `summarypoints` that contains two elements, the spatial information and the summaries:
```{r}
names(res_sum)
```
The summary for the first stand can now be accessed through the first element of `summarylist`:
```{r}
res_sum$summarylist[[1]]
```

Summary objects are handy because they allow us to display maps:
```{r, fig.align='center', fig.width=7}
plot(res_sum, "Transpiration", "2001-01-01", limits=c(0,35))
plot(res_sum, "Transpiration", "2001-02-01", limits=c(0,35))
```


### Simulation with integrated summaries

If one needs to save memory, it is possible with `spwbpoints()` to generate summaries automatically after the simulation of soil water balance of each stand, and storing those summaries instead of all the output of `spwb()`. For example the following code will keep summaries instead of simulation results: 

```{r, warning=FALSE}
res_2 <- spwbpoints(examplepointslandscape, SpParamsMED, examplemeteo, 
                  dates = dates, localControl = localControl,                  
                  keepResults = FALSE, progress = FALSE,
                  summaryFunction = summary.spwb, summaryArgs = list(freq="months"))
```

Now the summaries are embedded in the simulation results, and can be directly accessed through the element `summarylist`:
```{r}
res_2$summarylist[[1]]
```

And one can produce plots directly from simulation results:
```{r, fig.align='center', fig.width=7}
plot(res_2, "Transpiration", "2001-02-01")
```

## Simulation of watershed hydrology

Simulation of watershed hydrology involves describing a watershed in a gridded mode and simulating soil water balance water while including additional compartments and hydrological processes, such as routing runoff from one cell to the other. 

### Example data set
To illustrate this kind of simulation, which is carried out using function `spwbland()`, we first load a small example watershed included with the package:

```{r, warning=FALSE, fig.align="center", fig.width=7}
data("examplewatershed")
plot(examplewatershed, "elevation")
```

### Simulation function
As for the call to function `spwbpoints()` we will use the same weather (i.e. `examplemeteo`) across the watershed. To speed up calculations we call function `spwbland()` for a single month.

```{r, warning=FALSE}
dates = seq(as.Date("2001-01-01"), as.Date("2001-01-31"), by="day")
res_ws = spwbland(examplewatershed, SpParamsMED, examplemeteo, dates = dates, summaryFreq = "month")
```

Although simulations are performed using daily temporal steps, parameter `summaryFreq` allows storing results at coarser temporal scales, to reduce the amount of memory in spatial results. 

### Summaries and plots
Unlike `spwbpoints()` where summaries could be arbitrarily generated *a posteriori* from simulation results, with `spwbland()` the summaries are always fixed and embedded with the simulation result. For example, we can inspect the summaries for a given landscape cell using:
```{r}
res_ws$summarylist[[1]]
```

Several plots can be drawn from the result of function `spwbland()` in a similar way as done for `spwbpoints()`. As an example we display a map of soil water volume for the simulated month:
```{r, warning=FALSE, fig.align='center', fig.width=7}
plot(res_ws, "SoilVol", "2001-01-01")
```
