---
title: "Landscape water balance"
author: "Miquel De Caceres"
date: "`r Sys.Date()`"
output: 
  rmarkdown::html_vignette:
    toc: TRUE
vignette: >
  %\VignetteIndexEntry{Landscape water balance}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignettePackage{medfateland}
  \usepackage[utf8]{inputenc}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(medfate)
library(medfateland)
```


## Displaying properties of spatial objects

We begin by loading an example dataset of 100 forest stands distributed on points in the landscape:

```{r, fig.align='center', fig.width=7}
data("examplepointslandscape")
```

Using `plot()` functions for spatial landscape objects, we can draw maps of some variables using:

```{r, fig.align='center', fig.width=7}
plot(examplepointslandscape, "basalArea")
```

The set of maps available can be known by inspecting the help of function `getLandscapeLayer()`. Alternatively, the package provides function `shinyplotland()` to display maps interactively.

## Simulations on a set of forest stands

Package `medfate` includes functions `spwb()`, `growth()` and `fordyn()` to simulate soil water balance, carbon balance and forest dynamics on a single forest stand, respectively. This section describe how to run simulations on a set of forest stands in one call. This is done using functions `spwbspatial()`, `growthspatial()` and `fordynspatial()`, respectively. 

As an example, we will use function `spwbspatial()`, which simulates soil plant water balance on forests distributed in particular locations. The function takes an object of class `Spatial_*_Landscape` as input.

### Weather inputs and local control parameters

As before, we need meteorological data, species parameters and control parameters for local simulations (which will apply to all forest stands):

```{r}
data("examplemeteo")
data("SpParamsMED")
localControl = defaultControl()
```

### Calling the simulation function

The call to `spwbspatial()` can be done as follows (here we restrict the dates for simplicity):
```{r, warning=FALSE}
dates = seq(as.Date("2001-01-01"), as.Date("2001-02-28"), by="day")
res <- spwbspatial(examplepointslandscape, SpParamsMED, examplemeteo, 
                  dates = dates, localControl = localControl,  progress = FALSE)
```

Function `spwbspatial()` first initializes model inputs by calling `forest2spwbInput()` for each forest stand described in the `Spatial_*_Landscape` object.  Then it calls function `spwb()` for each forest stand and stores the result. The fact that we used `examplemeteo` as meteorological input involves that the same weather was applied to all forest stands, but different weather could have been specified for each one (see documentation of function `spwbspatial()`).

The result of calling `spwbspatial()` is a list (whose class name is the same as the function name) with the following elements:

```{r}
names(res)
```

Element `sp` describes spatial information, vector `xlist` contains the `spwbInput` corresponding to each forest stand (which can be used in subsequent simulations), vector `resultlist` contains the output of `spwb()` function for each forest stand (i.e. its elements are objects of the S3 class `spwb`) and `summarylist` contains summaries (if requested, see below). 

### Temporal summaries, plots and maps

The structure of the output of `spwbspatial()` allows querying information for the simulation of any particular forest stand. For example, we may use function `plot.spwb()`, from package **medfate**, to display the simulation results on a particular plot:
```{r, fig.width=7}
plot(res$resultlist[[1]], "Evapotranspiration")
```

Similarly, if we want a monthly summary of water balance for the first stand, we can use function `summary.spwb()` from package **medfate**:

```{r}
summary(res$resultlist[[1]], freq="months",FUN=sum, 
        output="WaterBalance")
```

However, a more convenient way of generating summaries is by calculating them on all forest stands in one step, using function `summary()` for objects of class `spwbspatial`:

```{r}
res_sum <- summary(res, summaryFunction = summary.spwb, 
                   freq="months", output="WaterBalance")
```
where the argument `summaryFunction` points to the function to be used to generate local summaries and the remaining arguments are those of the local summary function. The result of using `summary()` on an object `spwbspatial` is an object of class `summaryspatial` that contains two elements, the spatial information and the list of summaries for all stands:
```{r}
names(res_sum)
```
The summary for the first stand can now be accessed through the first element of `summarylist`:
```{r}
res_sum$summarylist[[1]]
```

Objects of class `summaryspatial` are handy because their `plot()` function allows us to display maps of temporal summaries:
```{r, fig.align='center', fig.width=7}
plot(res_sum, "Transpiration", "2001-01-01", limits=c(0,45))
plot(res_sum, "Transpiration", "2001-02-01", limits=c(0,45))
```

To avoid displaying maps one by one, the package includes function `shinyplotland()` that allows displaying maps of temporal summaries interactively.

## Advanced simulation features

### Simulation with integrated temporal summaries

If one needs to save memory, it is possible with `spwbspatial()` to generate temporal summaries automatically after the simulation of soil water balance of each stand, and storing those summaries instead of all the output of function `spwb()`. 

For example the following code will keep temporal summaries of water balance components instead of simulation results: 

```{r, warning=FALSE}
res_2 <- spwbspatial(examplepointslandscape, SpParamsMED, examplemeteo, 
                  dates = dates, localControl = localControl,                  
                  keepResults = FALSE, progress = FALSE,
                  summaryFunction = summary.spwb, summaryArgs = list(freq="months"))
```

Parameter `keepResults = FALSE` tells `spwbspatial()` not to keep the simulation results of forest stands, whereas `summaryFunction = summary.spwb` tells `spwbspatial()` to perform and store summaries before discarding the results of any stand. 

Now the temporal summaries are embedded in the simulation results, and can be directly accessed through the element `summarylist`:
```{r}
res_2$summarylist[[1]]
```

And one can produce maps with summary results directly from the output of the simulation function:
```{r, fig.align='center', fig.width=7}
plot(res_2, "Transpiration", "2001-02-01", limits=c(0,45))
```

### Simulation in several steps

The result of a simulation includes an element `xlist`, which stores the state of soil and stand variables at the end of the simiulation. This information can be used to perform a new simulation from the point where the first one ended. In order to do so, we need to update the state variables in spatial object with their values at the end of the simulation, using function `updateState()`:

```{r}
y = updateState(examplepointslandscape, res)
```

The resulting object `y` is the same as `examplepointslandscape` except for the state variables. For example we can compare the water potential in the first soil layer:

```{r, fig.align='center', fig.width=7}
plot(examplepointslandscape, "psi1")
plot(y, "psi1")
```

By using `y` as input we can now simulate water balance in the set of stands for an extra month:

```{r, warning=FALSE}
dates = seq(as.Date("2001-03-01"), as.Date("2001-03-31"), by="day")
res_3 <- spwbspatial(y, SpParamsMED, examplemeteo, 
                    dates = dates, localControl = localControl, 
                    summaryFunction = summary.spwb, summaryArgs = list(freq="months"),
                    progress = FALSE)
```

And display a map with the resulting month transpiration:

```{r, fig.align='center', fig.width=7}
plot(res_3, "Transpiration", "2001-03-01", limits=c(0,45))
```

## Simulation of watershed hydrology

Simulation of watershed hydrology involves describing a watershed in a gridded mode and simulating soil water balance water while including additional compartments and hydrological processes, such as routing runoff from one cell to the other. 

### Example data set
To illustrate this kind of simulation, which is carried out using function `spwbland()`, we first load a small example watershed included with the package:

```{r, warning=FALSE, fig.align="center", fig.width=7}
data("examplewatershed")
plot(examplewatershed, "elevation")
```

### Simulation function
As for the call to function `spwbspatial()` we will use the same weather (i.e. `examplemeteo`) across the watershed. To speed up calculations we call function `spwbland()` for a single month.

```{r, warning=FALSE}
dates = seq(as.Date("2001-01-01"), as.Date("2001-01-31"), by="day")
res_ws = spwbland(examplewatershed, SpParamsMED, examplemeteo, dates = dates, summaryFreq = "month")
```

Although simulations are performed using daily temporal steps, parameter `summaryFreq` allows storing results at coarser temporal scales, to reduce the amount of memory in spatial results. 

### Summaries and plots
Unlike `spwbspatial()` where summaries could be arbitrarily generated *a posteriori* from simulation results, with `spwbland()` the summaries are always fixed and embedded with the simulation result. For example, we can inspect the summaries for a given landscape cell using:
```{r}
res_ws$summarylist[[1]]
```

Several plots can be drawn from the result of function `spwbland()` in a similar way as done for `spwbspatial()`. As an example we display a map of soil water volume for the simulated month:
```{r, warning=FALSE, fig.align='center', fig.width=7}
plot(res_ws, "SoilVol", "2001-01-01")
```
