---
title: "Landscape water balance"
author: "Miquel De Caceres"
date: "`r Sys.Date()`"
output: 
  rmarkdown::html_vignette:
    toc: TRUE
vignette: >
  %\VignetteIndexEntry{1. Landscape water balance}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignettePackage{medfateland}
  \usepackage[utf8]{inputenc}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(medfate)
library(medfateland)
```

## About this vignette

This document describes how to run the soil water balance model described in De Cáceres et al. (2015) over landscapes using package `medfateland`. 


### Water balance for a set of forest stands

So far, we used `spwb()` to simulate soil water balance on a single stand. Package `medfate` offers the possibility to conduct simulations on a set of forest stands in one call, using function `spwbpoints()`. This function takes an object of class `SpatialPointsForest` as input. We begin by loading an example dataset of 30 forest stands:

```{r}
data("exampleSPL")
plot(exampleSPL)
```

As before, we need meteorological data, species parameters and control parameters (which will apply to all forest stands):

```{r}
data("examplemeteo")
data("SpParamsMED")
control = defaultControl()
control$verbose = FALSE
```

The call to `spwbpoints()` can be done as follows:
```{r}
res <- spwbpoints(exampleSPL, SpParamsMED, examplemeteo, control = control)
```

For each forest stand described in the `SpatialPointsForest` object, function `spwbpoints()` first calls `forest2spwbInput()`, to generate the appropriate input, and then it calls `spwb()`. The elements of the result of calling `spwbpoints()` are:

```{r}
names(res)
```

where `sp` is object of class SpatialPoints with the spatial coordinates of forest stands. Vector `input` contains the `spwbInput` corresponding to each forest stand, and vector `result` contains the output of `spwb()` for each forest stand. This structure of the output of `spwbpoints()` allows querying information for the simulation of any forest stand. For example, if we want a monthly summary of species transpiration for the stand '80013', we can use:

```{r}
summary(res$result$`80013`, freq="months",FUN=sum, 
        output="PlantTranspiration", bySpecies = TRUE)
```

If one needs to save memory, it is possible with `spwbpoints()` to generate summaries automatically after the simulation of soil water balance of each stand, and storing those summaries instead of all the output of `spwb()`. 
