---
title: "Package overview"
author: "Miquel De Caceres"
date: "`r Sys.Date()`"
output: 
  rmarkdown::html_vignette:
    toc: TRUE
vignette: >
  %\VignetteIndexEntry{0. Package overview}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignettePackage{medfateland}
  \usepackage[utf8]{inputenc}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Introduction

The R package `medfate` has been designed to study the characteristics and simulate dynamics of Mediterranean forests. Fire and drought impacts are the main drivers covered by the package. Representation of vegetation is not spatially-explicit (i.e. trees or shrubs do not have explicit coordinates within forest stands). This simplified representation is chosen so that package functions can be easily applied to forest plot data from national forest inventories. Since the package intends to facilitate predictions of not only forest functioning but also forest structure and composition, the taxonomic identity of plants is stored, and parameter values need to be provided for each taxonomic entity (but the package could be used with functional groups). 

```{r}
library(medfateland)
```

## Data structures

### Forest objects

### Spatial objects
Package `medfateland` offers three spatial classes that inherit fields from three corresponding classes in package `meteoland`:

* `SpatialPointsLandscape`: represents a set of forest stands (including soil description) as points within a landscape. Extends class `SpatialPointsTopography`.
* `SpatialPixelsLandscape`: represents a set of forest stands (including soil description) or other land cover units (i.e. agricultural, rock outcrops or urban areas) as pixels within a gridded landscape. Extends class `SpatialPixelsTopography`.
* `SpatialGridLandscape`: represents a set of forest stands (including soil description) or other land cover units (i.e. agricultural, rock outcrops or urban areas) as pixels within a complete grid. Extends class `SpatialGridTopography`.

Spatial objects allow simulations to be performed more easily for a set of forest stands related spatially. Simulations on objects `SpatialPointsLandscape` are performed sequentially for all forest stands, but no landscape processes are considered. In contrast, simulating water balance on objects `SpatialGridLandscape` or `SpatialPixelsLandscape` allows considering spatial processes such as lateral water transfer.

## Dynamic simulation functions
### Water balance and plant drought stress

Eco-hydrological processes are fundamental for the simulation models included in the `medfate` package. Eco-hydrological processes can be divided into local and landscape processes. By local processes we refer to the water balance of soils and plants within forest stands. Processes affecting soil water content include rainfall, canopy interception, infiltration and runoff, percolation and deep drainage, soil evaporation and plant transpiration. In `medfate`, the soil water balance of a forest is primarily used to predict drought stress for living plants in it. Soil water balance can be studied for a single forest stand using function  `spwb()` or for multiple stands distributed spatially using function `spwbpoints()` or `spwbgrid()`. When input data is in form of continuous spatial variation of forests over a landscape, the `medfate` package allows studying local soil water balance in combination with water discharge from one cell to the other (a landscape hydrological process). This is done using function `spwbgrid()`. 


Functions `spwb()` and `spwbpoints()`  may be used to:

1. Monitor or forecast temporal variation in soil water content in particular stands (for example to estimate mushroom yield).
2. Monitor or forecast temporal variation of plant drought stress in particular stands (for example to anticipate mortality events).
3. Monitor or forecast temporal variation of fuel moisture in particular stands (for example to monitor wildfire risk).

Function `spwbgrid()` may be used to:

1. Determine areas where particular plant species have high vulnerability to suffer from drought stress.
2. Partition the total rainfall in given area among: 
    i) water evaporated from canopy interception or bare soil evaporation
    ii) water transpired by plants
    iii) water exported via runoff or deep drainage into the water table and river streams.

### Forest growth

Changes in leaf area and plant growth are key to evaluate the influence of climatic conditions on forest structure and function. Processes affecting annual changes leaf area and plant size are those involved water and carbon balances, as well as those affecting growth directly. Processes influencing plant water balance include those affecting soil water content, such as rainfall, canopy interception, infiltration and runoff, percolation and deep drainage, soil evaporation and plant transpiration. Carbon balance arises from the relationship between plant photosynthesis and respiration, although carbon reserves play a role in the availabity of carbon for growth. Water and carbon balances are coupled through the regulation of transpiration done by stomata. Plant growth is affected by the availability of carbon (source limitation), but also by temperature and water status (sink limitation). 

Package `medfate` allows simulating plant growth of a set of cohorts (competing for light and water) in a single forest stand using function `growth()` or for multiple stands distributed spatially using function `growthpoints()`. Leaf area of plants can be severely decreased by drought stress, causing an increase in dead leaf area and affecting carbon and water fluxes. Functions `growth()` and `growthpoints()` may be used to:

1. Monitor or forecast temporal variation in water fluxes and soil water content in particular stands (for example to estimate regulation ecosystem services) taking into account processes determining plant growth.
2. Monitor or forecast temporal variation of plant drought stress in particular stands(for example to anticipate mortality events) taking into account processes determining plant growth.
3. Monitor or forecast temporal variation of live fuel moisture and the amount of standing dead and live fuels in particular stands (for example to monitor wildfire risk).
