---
title: "Spatially-uncoupled simulations"
author: "Miquel De Caceres"
date: "`r Sys.Date()`"
output: 
  rmarkdown::html_vignette:
    toc: TRUE
vignette: >
  %\VignetteIndexEntry{Spatially-uncoupled simulations}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignettePackage{medfateland}
  \usepackage[utf8]{inputenc}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(medfate)
library(medfateland)
library(meteoland)
```

## Aim

The aim of this vignette is to illustrate how to use **medfateland** (v. `r packageVersion("medfateland")`) to carry out simulations of forest function and dynamics on a set of spatial units, without taking into account spatial processes. This is done using functions `spwb_spatial()`, `growth_spatial()` and `fordyn_spatial()`; which are counterparts of functions `spwb()`, `growth()` and `fordyn()` in package **medfate**. As an example, we will use function `spwb_spatial()`, but the other two functions would be used similarly.

## Preparation

### Input structures

The fundamental input structure in **medfateland** is an object of class **sf**, i.e. a simple feature collection where geometries (normally points) are described with attributes. In the case of **medfateland**, spatial attributes correspond to model inputs for each of the spatial units represented. We begin by loading an example data set of 100 forest stands distributed on points in the landscape:

```{r, fig.align='center', fig.width=7}
data("example_ifn")
example_ifn
```



Despite the geometries themselves (which have a coordinate reference system), the following columns are required:

  + `id` identifiers of stands (e.g. forest inventory plot codes)
  + `elevation` (in m), `slope` (in degrees), `aspect` (in degrees) describe the topography of the stands
  + `land_cover_type` describes the land cover in each unit (values should be *wildland* or *agriculture* for spatially-uncoupled simulations).
  + `forest` objects of class **forest** (see package **medfate**) describing the structure and composition of forest stands
  + `soil` describes the soil of each forest stand, using either data frame of physical attributes or initialized objects of class **soil** (see package **medfate**).

Note that columns `forest` and `soil` contain vectors of lists, where elements are either lists or data.frames. For example, the forest corresponding to the first stand is:

```{r}
example_ifn$forest[[3]]
```
and the soil is:

```{r}
example_ifn$soil[[3]]
```

### Displaying maps of landscape properties

Using `plot_variable()` functions for spatial landscape objects, we can draw maps of some variables using:

```{r, fig.align='center', fig.width=7}
plot_variable(example_ifn, "basal_area")
```

The set of maps available can be known by inspecting the help of function `extract_variables()`. Alternatively, the package provides function `shinyplot_land()` to display maps interactively.

### Climate forcing

In **medfateland** there are several alternatives to specify climate forcing. 

(1) Provide a single data frame with daily weather for all stands.
(2) Provide a different daily weather for each stand (in an additional column of `sf` called `meteo`)
(3) Provide an object to perform daily weather interpolation during simulations (see later in this document).

For most of this document, we will follow the simpler option, which is to assume the same weather for all stands. We will supply a single data frame with daily weather for all plots:

```{r}
data("examplemeteo")
head(examplemeteo)
```

More interesting alternatives are described in the help of function `spwb_spatial()`. Notably, a column `meteo` may be defined in the **sf** input object, where for each spatial unit the user can supply a different data frame.

### Species parameters and local control parameters

Since it builds on **medfate**, simulations using **medfateland** require species parameters and control parameters for local simulations:

```{r}
data("SpParamsMED")
local_control <- defaultControl()
```

Importantly, the same control parameters will apply to all spatial units of the **sf** object.

## Carrying out simulations

As you should already know, package **medfate** includes functions `spwb()`, `growth()` and `fordyn()` to simulate soil water balance, carbon balance and forest dynamics on a single forest stand, respectively. This section describe how to run simulations on a set of forest stands in one call. As an example, we will use function `spwb_spatial()`, which simulates soil plant water balance on forests distributed in particular locations, but `growth_spatial()` and `fordyn_spatial()` are very similar. 


### Calling the simulation function

The call to `spwb_spatial()` can be done as follows (here we use parameter `dates` restrict the simulation period to january and february):
```{r, warning=FALSE}
dates <- seq(as.Date("2001-01-01"), as.Date("2001-02-28"), by="day")
res <- spwb_spatial(example_ifn, SpParamsMED, examplemeteo, 
                    dates = dates, local_control = local_control,  
                    parallelize = TRUE)
```

Function `spwb_spatial()` first initializes model inputs by calling `forest2spwbInput()` for each forest stand described in the **sf** landscape object.  Then it calls function `spwb()` for each forest stand and stores the result. In this case, we asked for parallel computation via the parameter `parallelize = TRUE`. 

The simulation result is also an object of class **sf** with the following columns:

```{r}
names(res)
```

Column `geometry` contains the geometry given as input to simulations, column `id` contains the identification label of each stand, column `state` contains the `spwbInput` corresponding to each forest stand (which can be used in subsequent simulations) and column `result` contains the output of `spwb()` function for each forest stand (i.e. its elements are objects of the S3 class `spwb`). 

### Temporal summaries, plots and maps

The structure of the output of `spwb_spatial()` allows querying information for the simulation of any particular forest stand. For example, we may use function `plot.spwb()`, from package **medfate**, to display the simulation results on a particular plot:
```{r, fig.width=7}
plot(res$result[[1]], "Evapotranspiration")
```

Similarly, if we want a monthly summary of water balance for the first stand, we can use function `summary.spwb()` from package **medfate**:

```{r}
summary(res$result[[1]], freq="months",FUN=sum, output="WaterBalance")
```

However, a more convenient way of generating summaries is by calculating them on all forest stands in one step, using function `simulation_summary()` on objects issued from simulations:

```{r}
res_sum <- simulation_summary(res, summary_function = summary.spwb, 
                              freq="months", output="WaterBalance")
```
where the argument `summary_function` points to the function to be used to generate local summaries and the remaining arguments are those of the local summary function. The result of using `simulation_summary()` is again an object of class `sf` that contains the spatial geometry and the list of summaries for all stands:
```{r}
names(res_sum)
```
The summary for the first stand can now be accessed through the first element of column `summary`:
```{r}
res_sum$summary[[1]]
```

Summary objects are handy because their `plot_summary()` function allows us to display maps of summaries for specific dates:
```{r, fig.align='center', fig.width=7}
plot_summary(res_sum, "Transpiration", "2001-01-01", limits=c(0,45))
plot_summary(res_sum, "Transpiration", "2001-02-01", limits=c(0,45))
```

To avoid displaying maps one by one, the package includes function `shinyplot_land()` that allows displaying maps of temporal summaries interactively.

### Simulation with integrated temporal summaries

If one needs to save memory, it is possible with `spwb_spatial()` to generate temporal summaries automatically after the simulation of soil water balance of each stand, and storing those summaries instead of all the output of function `spwb()`. 

For example the following code will keep temporal summaries of water balance components instead of simulation results: 

```{r, warning=FALSE}
res_2 <- spwb_spatial(example_ifn, SpParamsMED, examplemeteo, 
                  dates = dates, local_control = local_control,                  
                  keep_results = FALSE, parallelize = TRUE,
                  summary_function = summary.spwb, summary_arguments = list(freq="months"))
```

Parameter `keep_results = FALSE` tells `spwb_spatial()` not to keep the simulation results of forest stands, whereas `summary_function = summary.spwb` tells `spwb_spatial()` to perform and store summaries before discarding the results of any stand. The output has slightly different column names:

```{r}
names(res_2)
```

In particular, `result` is not included. Now the temporal summaries can be directly accessed through the column `summary`:
```{r}
res_2$summary[[1]]
```

And one can produce maps with summary results directly from the output of the simulation function:
```{r, fig.align='center', fig.width=7}
plot_summary(res_2, "Transpiration", "2001-02-01", limits=c(0,45))
```

The possibility of running a summary function after the simulation of each stand is not limited to `summary.spwb()`. Users can define their own summary functions, provided the first argument is `object`, which will contain the result of the simulation (i.e., the result of calling `spwb()`, `growth()` or `fordyn()`). For example, the following function returns the data frame corresponding to plant drought stress:

```{r}
f_stress <- function(object, ...) {
  return(object$Plants$PlantStress)
}
```

Now we can call again `spwb_spatial`:
```{r, warning=FALSE}
res_3 <- spwb_spatial(example_ifn, SpParamsMED, examplemeteo,
                  dates = dates, local_control = local_control,
                  keep_results = FALSE, parallelize = TRUE,
                  summary_function = f_stress)
```

The drought stress summary of stand 3 is:
```{r}
head(res_3$summary[[3]])
```


### Continuing a previous simulation

The result of a simulation includes an element `state`, which stores the state of soil and stand variables at the end of the simulation. This information can be used to perform a new simulation from the point where the first one ended. In order to do so, we need to update the state variables in spatial object with their values at the end of the simulation, using function `update_landscape()`:

```{r}
example_ifn_mod <- update_landscape(example_ifn, res)
names(example_ifn_mod)
```

Note that a new column `state` appears in now in the **sf** object. We can check the effect by drawing the water potential in the first soil layer:

```{r, fig.align='center', fig.width=7}
plot_variable(example_ifn_mod, "psi1")
```

By using this new object as input we can now simulate water balance in the set of stands for an extra month:

```{r, warning=FALSE}
dates <- seq(as.Date("2001-03-01"), as.Date("2001-03-31"), by="day")
res_3 <- spwb_spatial(example_ifn_mod, SpParamsMED, examplemeteo, 
                      dates = dates, local_control = local_control, 
                      summary_function = summary.spwb, summary_arguments = list(freq = "months"),
                      parallelize = TRUE)
```

And display a map with the resulting month transpiration:

```{r, fig.align='center', fig.width=7}
plot_summary(res_3, "Transpiration", "2001-03-01", limits=c(0,45))
```

### Simulations using weather interpolation

Specifying a single weather data frame for all the forest stands may be not suitable if they are very far from each other. Specifying a different weather data frame for each stand can also be a problem if there are thousands of stands to simulate, due to the huge data requirements. A solution for this can be using interpolation on the fly, inside simulations. This can be done by supplying an **interpolator** object (or a list of them), as defined in package **meteoland**. Here we use the example data provided in the package:

```{r}
interpolator <- meteoland::with_meteo(meteoland_meteo_example, verbose = FALSE) |>
    meteoland::create_meteo_interpolator(params = defaultInterpolationParams())
```

Once we have this object, using it is straightforward, via parameter `meteo`:
```{r, warning=FALSE}
res_4 <- spwb_spatial(example_ifn_mod, SpParamsMED, meteo = interpolator, 
                      local_control = local_control, 
                      summary_function = summary.spwb, summary_arguments = list(freq = "months"),
                      parallelize = TRUE)
```

