---
title: "Watershed simulations"
author: "Miquel De Caceres"
date: "`r Sys.Date()`"
output: 
  rmarkdown::html_vignette:
    toc: TRUE
vignette: >
  %\VignetteIndexEntry{Watershed simulations}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignettePackage{medfateland}
  \usepackage[utf8]{inputenc}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(medfate)
library(medfateland)
library(meteoland)
```

## Aim

The aim of this vignette is to illustrate how to use **medfateland**  (v. `r packageVersion("medfateland")`) to carry out simulations of forest function and dynamics on a set of forest stands while including lateral water transfer processes. This is done using functions `spwb_land()`, `growth_land()` and `fordyn_land()`; which are counterparts of functions `spwb()`, `growth()` and `fordyn()` in package **medfate**. As an example, we will use function `spwb_land()`, but the other two functions would be used similarly.

## Preparation

Preparing inputs for watershed simulations can be tedious. Two main inputs need to be assembled, described in the following two sections.

### Input sf objects

Here we load a small example watershed included with the package, that can be used to understand the inputs required:

```{r, warning=FALSE}
data("example_watershed")
example_watershed
```

Some of the columns like `forest`, `soil`, `elevation`, or `state`, were also present in the example for spatially-uncoupled simulations, so we will not repeat them. The following describes additional columns that are relevant here.

**Land cover type**

Simulations over watersheds normally include different land cover types. These are described in column `land_cover_type`:

```{r}
table(example_watershed$land_cover_type)
```

Local and landscape processes will behave differently depending on the land cover type. 

**Aquifer and snowpack**

Columns `aquifer` and `snowpack` are used as state variables to store the water content in the aquifer and snowpack, respectively.

**Crop factors**

Since the landscape contains agricultural lands, we need to define *crop factors*, which will determine transpiration flow as a proportion of potential evapotranspiration:

```{r}
example_watershed$crop_factor = NA
example_watershed$crop_factor[example_watershed$land_cover_type=="agriculture"] = 0.75
```


### Grid topology

Note that the `sf` structure does not imply a grid *per se*. Point geometry is used to describe the central coordinates of grid cells, but does not describe the grid. This means that another spatial input is needed to describe the grid topology, which in our case is an object of class `SpatRaster` from package **terra**:

```{r}
r <-terra::rast(xmin = 401380, ymin = 4671820, xmax = 402880, ymax = 4672620, 
                nrow = 8, ncol = 15, crs = "epsg:32631")
r
```

The `r` object must have the same coordinate reference system as the `sf` object. Moreover, each grid cell can contain up to one point of the `sf` (typically at the cell center). Some grid cells may be empty, though, so that the actual simulations may be done on an incomplete grid. Note that the raster does not contain data, only the topology is needed (to define neighbors and cell sizes, for example). All relevant attribute data is already included in the `sf` object.

### Watershed control options

Analogously to local-scale simulations with **medfate**, watershed simulations have overall control parameters. Notably, the user needs to decide which sub-model will be used for lateral water transfer processes (a decision similar to choosing the plant transpiration sub-model in **medfate**), by default "tetis":

```{r}
ws_control <- default_watershed_control("tetis")
```


## Carrying out simulations

### Launching watershed simulations

To speed up calculations, we call function `spwb_land()` for a single month:

```{r, warning=FALSE}
dates <- seq(as.Date("2001-01-01"), as.Date("2001-01-31"), by="day")
res_ws1 <- spwb_land(r, example_watershed,
                    SpParamsMED, examplemeteo, dates = dates, summary_frequency = "month",
                    watershed_control = ws_control)
```

Although simulations are performed using daily temporal steps, parameter `summary_frequency` allows storing results at coarser temporal scales, to reduce the amount of memory in spatial results. 

### Structure of simulation outputs

Function `spwb_land()` and `growth_land()` return a list with the following elements:

```{r}
names(res_ws1)
```

Where `sf` is an object of class `sf`, analogous to those of functions `*_spatial()`:

```{r}
res_ws1$sf
```
Columns `state`, `aquifer` and `snowpack` contain state variables, whereas `summary` contains temporal water balance summaries for all cells. Column `result` is empty in this case, but see below.

The other elements of the simulation result list (`watershed_balance`, `watershed_soil_balance` and `daily_export_runoff_m3`) refer to watershed level results.

### Accessing and plotting cell summaries

Unlike `spwb_spatial()` where summaries could be arbitrarily generated *a posteriori* from simulation results, with `spwb_land()` the summaries are always fixed and embedded with the simulation result. For example, we can inspect the summaries for a given landscape cell using:
```{r}
res_ws1$sf$summary[[1]]
```

Several plots can be drawn from the result of function `spwb_land()` in a similar way as done for `spwb_spatial()`. As an example we display a map of soil water volume for the simulated month:

```{r, warning=FALSE, fig.align='center', fig.width=7}
plot_summary(res_ws1$sf, "SoilVol", "2001-01-01")
```

### Full simulation results for specific cells

The idea of generating summaries arises from the fact that local models can produce a large amount of results, of which only some are of interest at the landscape level. Nevertheless, it is possible to specify those cells for which full daily results are desired. This is done by adding a column `result_cell` in the input `sf` object:

```{r}
# Set request for daily model results in cells number 3 and 9
example_watershed$result_cell <- FALSE
example_watershed$result_cell[c(3,9)] <- TRUE
```

If we launch the simulations again (omitting progress information):

```{r}
res_ws1 <- spwb_land(r, example_watershed,
                    SpParamsMED, examplemeteo, dates = dates, summary_frequency = "month",
                    watershed_control = ws_control, progress = FALSE)
```

We can now retrieve the results of the desired cell, e.g. the third one, in column `result` of `sf`:

```{r}
S <- res_ws1$sf$result[[3]]
class(S)
```

This object has class `spwb` and the same structure returned by function `spwb()` of **medfate**. Hence, we can inspect daily results using functions `shinyplot()` or `plot()`, for example:

```{r, warning=FALSE, fig.align='center', fig.width=7}
plot(S)
```

### Continuing a previous simulation

The result of a simulation includes an element `state`, which stores the state of soil and stand variables at the end of the simulation. This information can be used to perform a new simulation from the point where the first one ended. In order to do so, we need to update the state variables in spatial object with their values at the end of the simulation, using function `update_landscape()`:

```{r}
example_watershed_mod <- update_landscape(example_watershed, res_ws1)
names(example_watershed_mod)
```

Note that a new column `state` appears in now in the **sf** object. We can check the effect by drawing the water potential in the first soil layer:

```{r, fig.align='center', fig.width=7}
plot_variable(example_watershed_mod, "psi1")
```

Now we can continue our simulation, in this case adding an extra month:

```{r, warning=FALSE}
dates <- seq(as.Date("2001-02-01"), as.Date("2001-02-28"), by="day")
res_ws3 <- spwb_land(r, example_watershed_mod,
                    SpParamsMED, examplemeteo, dates = dates, summary_frequency = "month",
                    watershed_control = ws_control)
```

The fact that no cell required initialization is an indication that we used an already initialized landscape.


### Simulations of watershed forest dynamics

Running `growth_land()` is very similar to running `spwb_land()`. However, a few things change when we want to simulate forest dynamics using `fordyn_land()`. Regarding the `sf` input, an additional column `management_arguments` may be defined to specify the forest management arguments (i.e. silviculture) of cells. Furthermore, the function does not allow choosing the temporal scale of summaries. A call to `fordyn_land()` for a single year is given here (we subset the input to speed-up calculations):

```{r}
res_ws2 <- fordyn_land(r, example_watershed[1:10, ],
                       SpParamsMED, examplemeteo,
                       watershed_control = ws_control, progress = TRUE)
```

In this case, the `sf` part of the output contains additional columns, analogous to those of `fordyn_scenario()`.

```{r}
names(res_ws2$sf)
```



