---
title: "Evaluation of watershed-level runoff with river gauge data"
author: "Miquel De Cáceres / María González"
date: "`r Sys.Date()`"
description: >
   Benchmark evaluation at watershed level using river gauge data
output: 
  rmarkdown::html_document:
    df_print: paged
    toc: TRUE
---

```{r setup, include=FALSE}
library(kableExtra)
library(sf)
library(knitr)
library(medfate)
library(ggplot2)
library(cowplot)
library(hydroGOF)
evalDIR = "~/OneDrive/mcaceres_work/model_development/medfate_evaluation/WatershedLevelEvaluation/"
```

```{r, echo = FALSE}
medfateland_versions <- list.files(file.path(evalDIR, "data"))
medfateland_versions<-medfateland_versions[!(medfateland_versions %in% "observed")]
version_vec <- character(0)
model_vec <- character(0)
watershed_vec <- character(0)
for(medfateland_version in medfateland_versions) {
  models <- list.files(file.path(evalDIR, "data", medfateland_version))
  for(model in models) {
    watersheds <- list.files(file.path(evalDIR, "data", medfateland_version, model))
    for(watershed in watersheds) {
      version_vec <- c(version_vec, medfateland_version)
      model_vec <- c(model_vec, models)
      watershed_vec <- c(watershed_vec, watershed)
    }
  }
}
df <- tibble::tibble(version = version_vec, model = model_vec, watershed = watershed_vec) 
df$before <- vector("list", nrow(df))
df$after <- vector("list", nrow(df))
df$Q_daily <- vector("list", nrow(df))
df$Q_monthly <- vector("list", nrow(df))
for(r in nrow(df)) {
  df$before[[r]] <- yaml::read_yaml(file.path(evalDIR, "data", df$version[r], df$model[r], df$watershed[r], "uncalibrated.yaml"))
  df$after[[r]] <- yaml::read_yaml(file.path(evalDIR, "data", df$version[r], df$model[r], df$watershed[r], "calibrated.yaml"))
  df$Q_daily[[r]] <- read.csv2(file.path(evalDIR, "data", df$version[r], df$model[r], df$watershed[r], "Q_daily.csv"))
  df$Q_monthly[[r]] <- read.csv2(file.path(evalDIR, "data", df$version[r], df$model[r], df$watershed[r], "Q_monthly.csv"))
}
```

## Introduction

### Goal

The aim of this article is to provide an assessment of the performance of `spwb_land()` for the prediction of watershed outflow. To this aim, we simulate hydrological processes in a set of benchmark watersheds and compare the model predictions of watershed outflow against measurements obtained using river gauges at watershed outlets. 

### Simulation procedure

For each watershed, the following procedure has been conducted:

  1. Initial warm-up simulation for a specified number of years
  2. Simulation for the period with observed data **before** calibrating watershed parameters
  3. Manual calibration of watershed parameters (to be replaced with automatized calibration)
  4. Final simulation for the period with observed data **after** calibrating watershed parameters.


### Goodness-of-fit statistics

The following goodness of fit statistics are calculated using package `hydroGOF`:

+ Nash-Sutcliffe Efficiency (NSE): This coefficient is sensitive to extreme values and might yield sub-optimal results when the dataset contains large outliers.
+ Kling–Gupta efficiency (KGE): Provides a decomposition of the Nash-Sutcliffe efficiency, which facilitates the analysis of the importance of different components (bias, correlation and variability).
+ Index of agreement (d): Initially proposed by Willmott (1981) to overcome the drawbacks of the R2, such as the differences in observed and predicted means and variances (Legates and McCabe, 1999). d is also dimensionless and bounded between 0 and 1 and can be interpreted similarly to R2.
+ Volumetric efficiency index (VE): Originally proposed by Criss and Winston (2008) to circumvent some of the NSE flaws. VE values are also bounded [0, 1] and represent the fraction of water delivered at the proper time

### Hydrological analysis


## Watershed (TETIS) parameters

The following table contains the set of TETIS parameters employed in `spwb_land()` simulations on all watersheds, before and after calibration:

```{r, echo = FALSE}
df_params <- df |> 
  dplyr::select(-Q_daily, -Q_monthly) |>
  tidyr::pivot_longer(cols = c(before, after), names_to = "Calibration", values_to ="Params")
df_params$R_localflow <- NA
df_params$R_interflow <- NA
df_params$R_baseflow <- NA
df_params$n_interflow <- NA
df_params$n_baseflow <- NA
df_params$num_daily_substeps <- NA
df_params$rock_max_infiltration <- NA
df_params$deep_aquifer_loss <- NA
for(i in 1:nrow(df_params)) {
  df_params$R_localflow[i] <- df_params$Params[[i]]$tetis_parameters$R_localflow
  df_params$R_interflow[i] <- df_params$Params[[i]]$tetis_parameters$R_interflow
  df_params$R_baseflow[i] <- df_params$Params[[i]]$tetis_parameters$R_baseflow
  df_params$n_interflow[i] <- df_params$Params[[i]]$tetis_parameters$n_interflow
  df_params$n_baseflow[i] <- df_params$Params[[i]]$tetis_parameters$n_baseflow
  df_params$num_daily_substeps[i] <- df_params$Params[[i]]$tetis_parameters$num_daily_substeps
  df_params$rock_max_infiltration[i] <- df_params$Params[[i]]$tetis_parameters$rock_max_infiltration
  df_params$deep_aquifer_loss[i] <- df_params$Params[[i]]$tetis_parameters$deep_aquifer_loss
}
df_params |>
  dplyr::select(-Params) |>
  kableExtra::kbl() |>
  kableExtra::kable_styling()
```


## Evaluation results

```{r, results='asis', warning = FALSE, message=FALSE, echo = FALSE}
for(i in 1:nrow(df)) {
   medfateland_version <- df$version[i] 
   model <- df$model[i]
   watershed <- df$watershed[i]
   Q_daily <- df$Q_daily[[i]] |>
    dplyr::mutate(Date = as.Date(Date))
  Q_monthly <- df$Q_monthly[[i]] |>
    dplyr::mutate(Date = as.Date(Date))
  res <- knitr::knit_child('_WLEvaluation_child.Rmd', quiet = TRUE)
  cat(res, sep = '\n')
}
```



TO BE DONE

